Eliminating prior-bias from sparse-projection tomographic reconstructions

### 2020-07-11 paper/tci.tex
very minor edits

Most object-prior based reconstruction has been in the areas of
dynamic CT and 4D-CT. In dynamic CT, the object being scanned
undergoes changes between successive projection views and hence is not
stationary even within a single complete scan. In contrast, in 4D-CT,
a set of 3D CT scans are taken over short intervals of time, and the
object undergoes changes between successive scans.  The reconstruction
is then performed over the entire set generating several 3D
volumes. ~\cite{PICCS} is one of the earliest work to establish the
use of object-prior along with sparsity prior in dynamic CT, in an
attempt to reduce the number of views. The limitation is the
unchecked over-emphasis of the prior on the reconstruction of the test
at hand. This limitation was later dealt with by applying very
specific object-properties for reconstruction. For example,
in~\cite{Van2015}, the knowledge of the attenuation coefficient of the
fluid is used as a prior for reconstructing its flow through
gravel. In~\cite{Koen2020}, the changes across the sucessive scan
volumes is assumed to be continuous, thus enabling the use of optical
flow to model the motion between corresponding voxels of different
scan volumes. In another study~\cite{vincent2017}, changes between
successive scans were modelled by affine deformation whose parameters
were computed for motion-correction. Such a correction enables
acquisition from fewer views than would have otherwise not been
possible for reconstruction with adequate
fidelity. In~\cite{daniil2015}, all the scan volumes are reconstructed
together using spatio-temporal regularization.

@@

%---------------------------------Uniform section---------------------
\section{Method: uniform prior-based reconstruction}
\label{sec:method_uniform_prior}
Having presented the application, we first review the algorithm~\cite{my_dicta_paper} for a uniform (global) prior-based reconstruction. Principal Component Analysis (PCA) has been traditionally used to find the significant modes of (Gaussian corrupted) data. In this regard, it has been widely applied in the context of data compression. However, PCA can also be seen as a tool to provide an orthogonal basis to represent the space in which the test data could lie. This space is constructed from the available set of previously scanned objects which must cover a realistically representative range of structures. We first present the eigenspace-cum-CS prior-based reconstruction, which was shown~\cite{my_dicta_paper} to be better when compared to dictionary-based priors. %A preliminary version of this particular method was earlier presented in \cite{my_dicta_paper}.

To begin with, when an object is scanned multiple times, a set of high quality reconstructions (i.e. reconstructions from a dense set of projection views) may be chosen as object-prior for the reconstruction of future scan volumes, which in turn, may be scanned using far fewer measurements. The eigenspace $E_{\text{high}}$ of the $L$ previously scanned objects $Q_1,Q_2,...,Q_L$ is pre-computed. Here, it is assumed that the test volume can be expressed as a sparse linear combination of the principal components (eigenvectors of the covariance matrix) obtained from a group of structurally similar volumes. Hence, the object-prior is represented by means of PCA. For the eigenspace to encompass a range of possible structures in the test slice, the object-prior must represent a wide structural range. Moreover, if these volumes are not aligned, then they must be first registered before computing the prior. The prior is built by computing the covariance matrix from the template set $\{Q_i\}_{i=1}^L$. \emph{The space spanned by the eigenvectors $\{\boldsymbol{V_k}\}_{k=1}^{L-1}$ (eigenspace) of the covariance matrix is the global prior and is assumed to contain any test slice that is similar, but not necessarily identical to the object-prior.}  We use all of the $L-1$ orthogonal eigenvectors as a basis to represent the unknown test volume. Let $\boldsymbol{\mu}$ denote the mean of the previously scanned objects, and $\boldsymbol{\alpha}$ the vector of eigen-coefficients of the test scan, of which $\alpha_k$ is the $k^{\textrm{th}}$ element. Then, once the eigenspace is pre-computed, the test is reconstructed by minimizing the following cost function:
\vspace{4mm}
 \begin{equation}
   \setlength{\belowdisplayskip}{0pt} \setlength{\belowdisplayshortskip}{0pt}
\setlength{\abovedisplayskip}{-2pt} \setlength{\abovedisplayshortskip}{-2pt}
J_1(\boldsymbol{\theta,\alpha}) = \lVert\boldsymbol{\mathcal{R} x}- \boldsymbol{y}\rVert_2^2  + \lambda_1\lVert\boldsymbol{\theta}\rVert_1+\lambda_2\lVert\boldsymbol{x} - (\boldsymbol{\mu} + \sum_{k}\boldsymbol{V_k}\alpha_k)\rVert_2^2.
\label{Eq:main_prior}
\end{equation}
Here, $\lambda_1$, $\lambda_2$ are tunable weights given to the sparsity and prior terms respectively. The unknowns $\boldsymbol{\theta}$ and $\boldsymbol{\alpha}$
are solved by alternately minimizing $J_{\boldsymbol{\alpha}}(\boldsymbol{\theta})$ using a fixed $\boldsymbol{\alpha}$, and $J_{\boldsymbol\theta}(\boldsymbol{\alpha})$ using the resultant $\boldsymbol{\theta}$, where 
\begin{align}
J_{\boldsymbol{\alpha}}(\boldsymbol{\theta}) &\triangleq \lVert\boldsymbol{\mathcal{R} x- y}\rVert_2^2  + \lambda_1\lVert\boldsymbol{\theta}\rVert_1+\lambda_2\lVert\boldsymbol{x} - (\boldsymbol{\mu + V\alpha})\rVert_2^2, \\
J_{\boldsymbol\theta}(\boldsymbol{\alpha}) &\triangleq \lVert\boldsymbol{\Upsilon\theta} - (\boldsymbol{\mu + V\alpha})\rVert_2^2.
\end{align}
Note that $\boldsymbol{\theta}$ is solved for using the basis pursuit CS solver~\cite{l1ls}. Solving for $\boldsymbol{\alpha}$ leads to the closed form update:
\begin{equation}
\boldsymbol{\boldsymbol{\alpha}} = \boldsymbol{V}^T(\boldsymbol{\Upsilon \theta} -\boldsymbol{\mu}).
\end{equation}
 Optimal values of $\lambda_1, \lambda_2$ must be empirically chosen \textit{a priori}, based on the reconstructions of one of the template volumes (see also Sec.~\ref{sec:discussion}). The cost function described in Eq.~\ref{Eq:main_prior} is biconvex and the convergence of this optimization is guaranteed by the monotone convergence theorem.
 %---------------------------------------------------------
 \vspace{2mm}

@@

 The proposed method has been validated on new\footnote {These and our
   code will be made available to the community.} scans of biological
 specimens in a longitudinal setting. These datasets in the form of
 \emph{raw cone-beam projection measurements} were acquired from a lab
 at the Australian National University (ANU). %We emphasize that in
 most of the literature on tomographic reconstruction, the results are
 shown on reconstruction from projections \emph{simulated} from 3D
 volumes. This is because most CT scanners do not reveal the raw
 projections, and instead output only the full reconstructed
 volumes. Moreover, the process of conversion from the projections to
 the full volumes is proprietary. \emph{Departing from this, we
   demonstrate reconstruction results from raw projection data.}
In all figures in this section, `uniform prior' refers to optimizing Eq.~\ref{eq:spatially-varying_prior} with $\boldsymbol{W}(x,y,z)=1$.
%----------------------------------------------------------
%-------------------------------------------------
\subsubsection{\textbf{Okra dataset}}
\label{Sec:okra_uniform}
Our first dataset is that of an Okra specimen consisting of its five
scans (Fig.~\ref{fig:object-prior_test_okra}). The measurements
consisted of real cone-beam projections from 450 views, each of size
$336\times156$. The corresponding size of the reconstructed volume is
$338\times338\times123$. Prior to the first scan, two holes were
drilled on the surface of the specimen. This was followed by four
scans, each after introducing one new cut. The specimen was kept in
the same position throughout the acquisitions. In cases where such an
alignment is not present, all the template volumes must be pre-aligned
before computing the eigenspace. The test must be registered to the
object-prior after its preliminary pilot reconstruction. The ground
truth consists of FDK reconstructed volumes from the the full set of
450 view projections. The test volume was reconstructed from a partial
set of 45 projections, i.e. $10\%$ of the projection views from which
ground truth was reconstructed. The selected 3D ground truth of
template volumes, the test volume as well as the 3D reconstructions
can be seen in~\cite{supp_paper}. One of the slices of the
reconstructed volumes is shown in Fig.~\ref{fig:okra_3D_results}. The
red and green 3D RoI in the video and images show the regions where
new changes are present. Based on the potato and the okra experiments,
we see that our method is able to \emph{discover both} the presence of
a new structure, as well as the absence of a structure.

@@

\section{Discussion}
\label{sec:discussion}
The presented uniform prior-based algorithm is suited to applications
where we wish to detect the location of new changes in an object over
time, and the presented spatially-varying prior-based algorithm is
suited to applications where we wish to observe the intricate details
of those new changes. This distinction in the techniques used has not
been studied earlier in
literature~\cite{geyer2015,PICCS,Ali2018,Tan2017} to the best of our
knowledge. Even though methods such as~\cite{Hakka2019} offer the flexibility of using multiple templates, the very important issue of these templates overwhelming the current reconstruction has been largely ignored. We have presented a principled method of tackling this particular issue via our weights map. We not only estimate the location of new changes, but also compute their intensities. This information is then used to modulate the strength of the prior during the reconstruction of the test. Our results consist of 3D reconstructions from real cone-beam projections of two datasets, and both 3D and 2D reconstructions from cone-beam and parallel beam projections respectively. In each case, we have compared our results with those from standard back-projection algorithms~\cite{Pan2009} (FDK in case of 3D and FBP incase of 2D), and Compressed Sensing~\cite{Donoho,introCS} based reconstructions. An in-depth analysis of each of the results is presented in Sec.~\ref{sec:results}. We now discuss the need for registration between object-prior and the test, the method that suits best when the data is homogeneous, and the effect of hyper-parameters on our reconstructions.

\begin{table}[!h]
  \centering
\caption{Computation time for reconstruction of volumes using various methods. The configuration of the system used was x86\_64 architecture, 62GB RAM, 12-core AMD 3.8GHz processor, with Nvidia GeForce RTX 2080 Ti GPU.}
\begin{tabular}{|l|c|c|c|c|c|c|c|c|}
\hline
   & \textbf{FDK} & \textbf{l1-ls} & \textbf{\begin{tabular}[c]{@{}c@{}}Uniform\\prior + l1-ls\end{tabular}} & \textbf{\begin{tabular}[c]{@{}c@{}}Spatially-Varying\\prior + l1-ls\end{tabular}} \\\hline
%\textbf{\begin{tabular}[c]{@{}c@{}}Potato\\size:(150,150,100)\\45 views\end{tabular}}  & 5 sec. & 8 min.  & 54 min. & 53 min. \\ \hline
\textbf{\begin{tabular}[c]{@{}c@{}}Okra\\size:(338,338,123)\\45 views\end{tabular}} & 6 sec. & 8 min.  & 25 min. & 60 min. \\ \hline
\textbf{\begin{tabular}[c]{@{}c@{}}Sprouts\\size:(130,130,130)\\xx views\end{tabular}}  &  &   &  &  \\ \hline
\end{tabular}
\label{table:Computation_time_3D}
\end{table}

>>>>

Most object-prior based reconstruction has been in the areas of
dynamic CT and 4D-CT. In dynamic CT, the object being scanned
undergoes changes between successive projection views and hence is not
stationary even within a single complete scan. In contrast, in 4D-CT,
a set of 3D CT scans are taken over short intervals of time, and the
object undergoes changes between successive scans.  The reconstruction
is then performed over the entire set generating several 3D
volumes. ~\cite{PICCS} is one of the earliest work to establish the
use of object-prior along with sparsity prior in dynamic CT, in an
attempt to reduce the number of views. The limitation is the
unchecked over-emphasis of the prior on the reconstruction of the test
at hand. This limitation was later dealt with by applying very
specific object-properties for reconstruction. For example,
in~\cite{Van2015}, the knowledge of the attenuation coefficient of the
fluid is used as a prior for reconstructing its flow through
gravel. In~\cite{Koen2020}, the changes across the sucessive scan
volumes are assumed to be continuous, thus enabling the use of optical
flow to model the motion between corresponding voxels of different
scan volumes. In another study~\cite{vincent2017}, changes between
successive scans were modelled by affine deformation whose parameters
were computed for motion-correction. Such a correction enables
acquisition from fewer views than would have otherwise not been
possible for reconstruction with adequate
fidelity. In~\cite{daniil2015}, all the scan volumes are reconstructed
together using spatio-temporal regularization.

@@

%---------------------------------Uniform section---------------------
\section{Method: uniform prior-based reconstruction}
\label{sec:method_uniform_prior}
Having presented the application, we first review the algorithm~\cite{my_dicta_paper} for a uniform (global) prior-based reconstruction. Principal Component Analysis (PCA) has been traditionally used to find the significant modes of (Gaussian corrupted) data. In this regard, it has been widely applied in the context of data compression. However, PCA can also be seen as a tool to provide an orthogonal basis to represent the space in which most of the test data could lie (except the new changes). This space is constructed from the available set of previously scanned objects which must cover a realistically representative range of structures. %, which was shown~\cite{my_dicta_paper} to be better when compared to dictionary-based priors. %A preliminary version of this particular method was earlier presented in \cite{my_dicta_paper}.

We first present the eigenspace-cum-CS prior-based reconstruction. To begin with, when an object is scanned multiple times, a set of high quality reconstructions (i.e. reconstructions from a dense set of projection views) may be chosen as object-prior for the reconstruction of future scan volumes, which in turn, may be scanned using far fewer measurements. The eigenspace $E_{\text{high}}$ of the $L$ previously scanned objects $Q_1,Q_2,...,Q_L$ is pre-computed. Here, it is assumed that the test volume can be expressed as a sparse linear combination of the principal components (eigenvectors of the covariance matrix) obtained from a group of structurally similar volumes. Hence, the object-prior is represented by means of PCA. For the eigenspace to encompass a range of possible structures in the test slice, the object-prior must represent a wide structural range. Moreover, if these volumes are not aligned, then they must be first registered before computing the prior. The prior is built by computing the covariance matrix from the template set $\{Q_i\}_{i=1}^L$. \emph{The space spanned by the eigenvectors $\{\boldsymbol{V_k}\}_{k=1}^{L-1}$ (eigenspace) of the covariance matrix is the global prior and is assumed to contain any test slice that is similar, but not necessarily identical to the object-prior.}  We use all of the $L-1$ orthogonal eigenvectors as a basis to represent the unknown test volume. Let $\boldsymbol{\mu}$ denote the mean of the previously scanned objects, and $\boldsymbol{\alpha}$ the vector of eigen-coefficients of the test scan, of which $\alpha_k$ is the $k^{\textrm{th}}$ element. Then, once the eigenspace is pre-computed, the test is reconstructed by minimizing the following cost function:
\vspace{4mm}
 \begin{equation}
   \setlength{\belowdisplayskip}{0pt} \setlength{\belowdisplayshortskip}{0pt}
\setlength{\abovedisplayskip}{-2pt} \setlength{\abovedisplayshortskip}{-2pt}
J_1(\boldsymbol{\theta,\alpha}) = \lVert\boldsymbol{\mathcal{R} x}- \boldsymbol{y}\rVert_2^2  + \lambda_1\lVert\boldsymbol{\theta}\rVert_1+\lambda_2\lVert\boldsymbol{x} - (\boldsymbol{\mu} + \sum_{k}\boldsymbol{V_k}\alpha_k)\rVert_2^2.
\label{Eq:main_prior}
\end{equation}
Here, $\lambda_1$, $\lambda_2$ are tunable weights given to the sparsity and prior terms respectively. The unknowns $\boldsymbol{\theta}$ and $\boldsymbol{\alpha}$
are solved by alternately minimizing $J_{\boldsymbol{\alpha}}(\boldsymbol{\theta})$ using a fixed $\boldsymbol{\alpha}$, and $J_{\boldsymbol\theta}(\boldsymbol{\alpha})$ using the resultant $\boldsymbol{\theta}$, where 
\begin{align}
J_{\boldsymbol{\alpha}}(\boldsymbol{\theta}) &\triangleq \lVert\boldsymbol{\mathcal{R} x- y}\rVert_2^2  + \lambda_1\lVert\boldsymbol{\theta}\rVert_1+\lambda_2\lVert\boldsymbol{x} - (\boldsymbol{\mu + V\alpha})\rVert_2^2, \\
J_{\boldsymbol\theta}(\boldsymbol{\alpha}) &\triangleq \lVert\boldsymbol{\Upsilon\theta} - (\boldsymbol{\mu + V\alpha})\rVert_2^2.
\end{align}
Note that $\boldsymbol{\theta}$ is solved for using the basis pursuit CS solver~\cite{l1ls}. Solving for $\boldsymbol{\alpha}$ leads to the closed form update:
\begin{equation}
\boldsymbol{\boldsymbol{\alpha}} = \boldsymbol{V}^T(\boldsymbol{\Upsilon \theta} -\boldsymbol{\mu}).
\end{equation}
 Optimal values of $\lambda_1, \lambda_2$ must be empirically chosen \textit{a priori}, based on the reconstructions of one of the template volumes (see also Sec.~\ref{sec:discussion}). The cost function described in Eq.~\ref{Eq:main_prior} is biconvex and the convergence of this optimization is guaranteed by the monotone convergence theorem.
 %---------------------------------------------------------
 \vspace{2mm}

@@

 The proposed method has been validated on new\footnote {These and our
   code will be made available to the community.} scans of biological
 specimens in a longitudinal setting. These datasets in the form of
 \emph{raw cone-beam projection measurements} were acquired from a lab
 at the Australian National University (ANU). Data from commercially
 available CT machines is not suitable because most CT scanners do
 not reveal the raw measurements, and instead output only the full
 reconstructed volumes. Moreover, the process of conversion from the
 projections to the full volumes is proprietary.
 %We emphasize that in
% most of the literature on tomographic reconstruction, the results are
 %shown on reconstruction from projections \emph{simulated} from 3D
 %volumes. This is because most CT scanners do not reveal the raw
 %projections, and instead output only the full reconstructed
 %volumes. Moreover, the process of conversion from the projections to
 %the full volumes is proprietary.
 \emph{Departing from this, we
   demonstrate reconstruction results from raw projection data.}
In all figures in this section, `uniform prior' refers to optimizing Eq.~\ref{eq:spatially-varying_prior} with $\boldsymbol{W}(x,y,z)=1$.
%----------------------------------------------------------
%-------------------------------------------------
\subsubsection{\textbf{Okra dataset}}
\label{Sec:okra_uniform}
Our first dataset is that of an Okra specimen consisting of its five
scans (Fig.~\ref{fig:object-prior_test_okra}). The measurements
consisted of real cone-beam projections from 450 views, each of size
$336\times156$. The corresponding size of the reconstructed volume is
$338\times338\times123$. Prior to the first scan, two holes were
drilled on the surface of the specimen. This was followed by four
scans, each after introducing one new cut. The specimen was kept in
the same position throughout the acquisitions. In cases where such an
alignment is not present, all the template volumes must be pre-aligned
before computing the eigenspace. The test must be registered to the
object-prior after its preliminary pilot reconstruction. The ground
truth consists of FDK reconstructed volumes from the the full set of
450 view projections. The test volume was reconstructed from a partial
set of 45 projections, i.e. $10\%$ of the projection views from which
ground truth was reconstructed. The selected 3D ground truth of
template volumes, the test volume as well as the 3D reconstructions
can be seen in~\cite{supp_paper}. One of the slices of the
reconstructed volumes is shown in Fig.~\ref{fig:okra_3D_results}. The
red and green 3D RoI in the video and images show the regions where
new changes are present. Based on the potato and the okra experiments,
we see that our method is able to \emph{discover both} the presence of
a new structure, as well as the absence of a structure.

@@

\section{Discussion}
\label{sec:discussion}
The presented uniform prior-based algorithm is suited to applications
where we wish to detect the location of new changes in an object over
time, and the presented spatially-varying prior-based algorithm is
suited to applications where we wish to observe the intricate details
of those new changes. This distinction in the techniques used has not
been studied earlier in
literature~\cite{geyer2015,PICCS,Ali2018,Tan2017} to the best of our
knowledge. Even though methods such as~\cite{Hakka2019} offer the flexibility of using multiple templates, the very important issue of these templates overwhelming the current reconstruction has been largely ignored. We have presented a principled method of tackling this particular issue via our weights map. We not only estimate the location of new changes, but also compute their intensities. This information is then used to modulate the strength of the prior during the reconstruction of the test. Our results consist of 3D reconstructions from real cone-beam projections of two datasets, and both 3D and 2D reconstructions from cone-beam and parallel beam projections respectively. In each case, we have compared our results with those from standard back-projection algorithms~\cite{Pan2009} (FDK in case of 3D and FBP incase of 2D), and Compressed Sensing~\cite{Donoho,introCS} based reconstructions. An in-depth analysis of each of the results is presented in Sections~\ref{sec:results_uniform_prior} and~\ref{sec:results_spatially_varying_prior}. We now discuss the need for registration between object-prior and the test, the method that suits best when the data is homogeneous, and the effect of hyper-parameters on our reconstructions.

\begin{table}[!h]
  \centering
\caption{Computation time for reconstruction of volumes using various methods. The configuration of the system used was x86\_64 architecture, 62GB RAM, 12-core AMD 3.8GHz processor, with Nvidia GeForce RTX 2080 Ti GPU.}
\begin{tabular}{|l|c|c|c|c|c|c|c|c|}
\hline
   & \textbf{FDK} & \textbf{l1-ls} & \textbf{\begin{tabular}[c]{@{}c@{}}Uniform\\prior + l1-ls\end{tabular}} & \textbf{\begin{tabular}[c]{@{}c@{}}Spatially-Varying\\prior + l1-ls\end{tabular}} \\\hline
%\textbf{\begin{tabular}[c]{@{}c@{}}Potato\\size:(150,150,100)\\45 views\end{tabular}}  & 5 sec. & 8 min.  & 54 min. & 53 min. \\ \hline
\textbf{\begin{tabular}[c]{@{}c@{}}Okra\\size:(338,338,123)\\45 views\end{tabular}} & 6 sec. & 8 min.  & 25 min. & 60 min. \\ \hline
\textbf{\begin{tabular}[c]{@{}c@{}}Sprouts\\size:(130,130,130)\\xx views\end{tabular}}  & 6 sec. & 3 hrs. & 60 min. &  \\ \hline
\end{tabular}
\label{table:Computation_time_3D}
\end{table}
